Informing policy via dynamic models: Cholera in Haiti

### 2024-01-31 ms.Rnw

\section*{Abstract}
Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
\eic{I'VE RE-WRITTEN HERE, LOOKING TO BE MORE SPECIFIC ABOUT THE GOALS AND CONTRIBUTION OF THE PAPER.}
\old{As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
We study three dynamic models developed by expert teams to advise on vaccination policies.
We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.}
\new{We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
We consider three dynamic models developed by expert teams to advise on vaccination policies.
We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
Specifically, we present approaches \old{to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{of diagnosing model misspecification, which facilitates the development of model improvements.
We further demonstrate the usefulness of recent advancements in likelihood optimization for high-dimensional nonlinear dynamic models, which enable improved consistency between high dimensional models and observed incidence data.}
\old{Our workflow is} \new{The presented case study is} reproducible and extendable, facilitating future investigations of this disease system.

>>>>

\section*{Abstract}
Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
\old{As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
We study three dynamic models developed by expert teams to advise on vaccination policies.
We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.}
\new{We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
We consider three dynamic models developed by expert teams to advise on vaccination policies.
We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
Specifically, we present approaches to diagnosis of model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models of spatiotemporal incidence data.}
Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
